### 2024-03-19 latex/HP3_comm.tex
new files and Nagari-Support for Jyotsna

\input{Preamble-Print}
\input{CommStyle}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{3.1}
\begin{translation}[hp03_001]
test.
\end{translation}

>>>>

\input{PreambleComm}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

%\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{3.1}
\begin{translation}[hp03_001]
Just as the lord of snakes is the foundation of the regions of the earth along with their mountains and forests, so \emph{kuṇḍalinī} is the foundation of all systems of yoga.
\end{translation}

%\begin{sources}[hp03_001]
%\end{sources}

\begin{testimonia}[hp03_001]
\emph{Haṭharatnāvalī} 2.124
\begin{versinnote}
\tl{saśailavanadhātryās tu yathādhāro 'hināyakaḥ |\\+}
\tl{aśeṣayogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~71v
\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{saśailavanadhātrīṇāṃ yathādhāro 'hināyakaḥ |\\+}
\tl{sarveṣāṃ yogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.170
\begin{versinnote}
\tl{tatra haṭhapradīpikāyām--\\+}
\tl{saśailavanadhātrīṇāṃ yathādhāro 'hināyakaḥ |\\+}
\tl{sarveṣāṃ yogatantrāṇāṃ tathādhāro hi kuṇḍalī ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_001]
The plural \emph{°dhātrīṇām} is hard to construe. Only this world has mountains and forests, but we want a plural for the comparison with \emph{°tantrāṇām}. Brahmānanda (\emph{Jyotsnā} 3.1) understands \emph{°dhātrīṇām} to refer to the different regions of the earth, even though the world (\emph{dhātrī}) is a single entity (\emph{dhātryā ekatve 'pi deśabhedād bhedam ādāya bahuvacanam}). A similar comment occurs in the \emph{Yogaprakāśikā} 5.1 (\emph{yathā samastadvīpādisahitapṛthvī ādhāraḥ phaṇīndras tathā samastayogādhāraḥ kuṇḍalīty āha saśaileti}). The author of the \textit{Haṭharatnāvalī} circumvented this issue by adopting the reading °\textit{dhātryāḥ}.
\end{philcomm}

%%%%%%%%%%
\subsection*{3.2}
\begin{translation}[hp03_002]
When the sleeping \emph{kuṇḍalinī} awakens through the favour of the guru, then all the lotuses are pierced, and the knots too, [\dots] 
\end{translation}

\begin{sources}[hp03_002]
\emph{Śivasaṃhitā} 4.21
\begin{versinnote}
\tl{suptā guruprasādena yadā jāgarti kuṇḍalī |\\+}
\tl{tadā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_002]
\emph{Yogacintāmaṇi} f.~71v (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{suptā guruprasādena bodhitā sukhadā bhavet |\\+}
\tl{tathā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 171 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{suptā guruprasādena yadā jāgarti kuṇḍalī |\\+}
\tl{tadā sarvāṇi padmāni bhidyante granthayo 'pi ca ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_002]
The usual meaning of \emph{jāgarti} would be “is wakeful” rather than “awakens”, which explains the variant \emph{bodhitā}.
\end{philcomm}

%%%%%%%%%%
\subsection*{3.3}
\begin{translation}[hp03_003]
[\dots] the empty pathway becomes the royal highway for \emph{prāṇa}, the mind becomes free of support, and death is cheated.
\end{translation}

\begin{sources}[hp03_003]
\end{sources}

\begin{testimonia}[hp03_003]
\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})

\begin{versinnote}
\tl{prāṇasya śūnyapadavī tathā rājapathāyate |\\+}
\tl{tathā cittaṃ nirālambaṃ tathā kālasya vañcanam ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp03_003]
%JM: take prāṇasya with śūnyapadavī?
%JB: tadā in 3b indicates that this verse should be read with 3.2 (i.e., [when kuṇḍalinī has awakened...] then, the empty pathway becomes the royal pathway for prāṇa)
%\end{philcomm}

\begin{metre}[hp03_003]
Anuṣṭubh (a: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{3.4}
What is ``the empty pathway"?

\begin{translation}[hp03_004]
\emph{Suṣumṇā}, the empty pathway, the great path to the aperture of Brahmā, the cremation ground, Śāmbhavī, and the middle path are synonyms.
\end{translation}

\begin{sources}[hp03_004]
Cf.~\emph{Amṛtasiddhi} 2.6
\begin{versinnote}
\tl{avadhūtīpadaṃ ke cic chmaśānaṃ ca mahāpatham |\\+}
\tl{ke cid vadanti ādhārāṃ suṣumnāṃ ca sarasvatīm ||\\!}
\end{versinnote}

Cf.~\emph{Dattātreyayogaśāstra} 109c–110b
\begin{versinnote}
\tl{mahāpathaṃ śmaśānaṃ ca suṣumnāpy ekam eva hi ||\\+}
\tl{nāmnāṃ matāntare bhedaḥ phale bhedo na vidyate |\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_004]
\emph{Yogacintāmaṇi} f.~59r
\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{suṣumṇā śūnyapadavī brahmarandhraṃ mahāpatham |\\+}
\tl{śmaśānī śāṃbhavī madhyamārgaś cety ekavācakā iti ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.172 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{prāṇasya śūnyapadavī mahārandhraṃ mahāpatham |\\+}
\tl{śmaśānaṃ śāmbhavī madhyamārgaś cety ekavācakam ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_004]
%In the context of \textit{suṣumṇā}, Svātmārāma seems to understand \emph{śāmbhavī} as the power (\emph{śakti}) of Śambhu (Cf.~\emph{Haṭhapradīpikā} 4.78c \emph{suṣumṇā śāmbhavī śaktiḥ}). JM: I'm not convinced by śāmbhavī = power of Śambhu; I think it's more feminine.
%\end{philcomm}

\begin{metre}[hp03_004]
Anuṣṭubh (a: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{3.5}
\begin{translation}[hp03_005]
Therefore, in order to do his utmost to awaken the goddess sleeping in front of the doorway of Brahman, [the yogi] should carry out the practice of \emph{mudrā}.
\end{translation}

\begin{sources}[hp03_005]
\emph{Śivasaṃhitā} 4.22
\begin{versinnote}
\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
\tl{brahmadvāramukhe suptāṃ mudrābhyāsaṃ samācaret ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_005]
\emph{Yogacintāmaṇi} f.~59r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
\tl{brahmadvāramukhe suptāṃ mudrābhyāsaparo bhavet ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.173 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tasmāt sarvaprayatnena prabodhayitum īśvarīm |\\+}
\tl{brahmadvāramukhe suptāṃ mudrābhyāsaṃ samācaret ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_005]
%Consider pathe for mukhe [JB: no support for pathe]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.6}
\begin{translation}[hp03_006]
The great seal, the great lock, the great piercing, the sky-roving [seal], the \emph{uḍḍiyāna} [lock], the root lock and the lock called \emph{jālandhara}, [\dots] 
\end{translation}

\begin{sources}[hp03_006]
%Cf.~Amṛtasiddhi 13.12
%\begin{versinnote}
%\tl{mahāmudrā mahābandho mahāvedhas tṛtīyakaḥ |\\!}
%\end{versinnote}

\emph{Śivasaṃhitā} 4.23
\begin{versinnote}
\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
\tl{jālandharo mūlabandho viparītakṛtis tathā ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_006]
\emph{Haṭharatnāvalī} 2.32

\begin{versinnote}
\tl{mahāmudrā mahābandho mahāvedhas tṛtīyakaḥ |\\+}
\tl{uḍḍiyānaṃ mūlabandho bandho jālandharābhidhaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
\tl{uḍḍiyānaṃ mūlabandho bandho jālandharābhidhaḥ ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.174 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{mahāmudrā mahābandho mahāvedhaś ca khecarī |\\+}
\tl{uḍyānaṃ mūlabandhaś ca bandho jālandharas tathā ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_006]
%JM: Do we translate mudrā or not?
%\end{philcomm}

\begin{metre}[hp03_006]
Anuṣṭubh (c: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{3.7}
\begin{translation}[hp03_007]
[\ldots] the bodily position called inverted, \emph{vajrolī} [and] the stimulation of the goddess: this group of ten \emph{mudrā}s and other [practices] destroys old age and death. 
\end{translation}

\begin{sources}[hp03_007]
\emph{Śivasaṃhitā} 4.24
\begin{versinnote}
\tl{uḍyānaṃ caiva vajrolī daśamaṃ śakticālanam |\\+}
\tl{idaṃ hi mudrādaśakaṃ mudrāṇām uttamottamam ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_007]
\emph{Haṭharatnāvalī} 2.33
\begin{versinnote}
\tl{karaṇī viparītākhyā vajrolī śakticālanam |\\+}
\tl{sampradāyā khecarī sā daśa mudrāḥ prakīrtitāḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{karaṇī viparītākhyā tathā vai śakticālanam |\\+}
\tl{etad dhi mudrānavakaṃ jarāmaraṇavarjitam ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.175 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{viparītakṛtiś caiva vajrolī śakticālanam |\\+}
\tl{idaṃ hi mudrādaśakaṃ mudrāṇām uttamomam ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp03_007]
%JB: hi (in 3.7c) is supported only by the \textit{Jyotsnā}. Mss. have tu and ca.
%\end{philcomm}

\begin{metre}[hp03_007]
Anuṣṭubh (c: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{3.8}
\begin{translation}[hp03_008]
It has been taught by Śiva, is divine, bestows the eight supramundane powers, is beloved of all the Siddhas, is difficult for even the gods to obtain, [\dots] 
\end{translation}

\begin{sources}[hp03_008]
\end{sources}

\begin{testimonia}[hp03_008]
\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ādināthoditaṃ divyam aṣṭaiśvaryapradāyakam |\\+}
\tl{vallabhaṃ sarvasiddhānāṃ durlabhaṃ marutām api ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.176 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ādināthoditaṃ samyagaṣṭaiśvaryapradāyakam |\\+}
\tl{vallabhaṃ sarvasiddhendradurlabhaṃ marutām api ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_008]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.9}
\begin{translation}[hp03_009]
[\ldots] should be carefully kept secret like a casket of gems [and] must not be spoken of to anyone, like sex with a respectable woman.
\end{translation}

\begin{sources}[hp03_009]
\end{sources}

\begin{testimonia}[hp03_009]
\emph{Yogacintāmaṇi} f.~72r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{gopanīyaṃ prayatnena yathā ratnakaraṇḍakam |\\+}
\tl{kasyacin naiva vaktavyaṃ kulastrīsurataṃ yathā ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.177 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{gopanīyaṃ prayatnena jarāmaraṇanāśanam |\\+}
\tl{kasyacinnaiva vaktavyaṃ kulastrīsurataṃ yathā || \\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_009]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.9*1}
\begin{translation}[hp03_009_1]
[With] \emph{Amarolī} and \emph{Sahajolī}, Vajrolī is considered to be threefold. I shall teach their characteristics and the details of how they should be performed.
\end{translation}

\begin{sources}[hp03_009_1]
\emph{Dattātreyayogaśāstra} 31c–32b
\begin{versinnote}
\tl{vajrolī cāmarolī ca sahajolī tridhā matā |\\+}
\tl{eteṣāṃ lakṣaṇaṃ vakṣye kartavyaṃ ca viśeṣataḥ ||\\+}
\tl{\var{vajrolī ] vajrolir \vl • cāmarolī ] amaroliś \vl • sahajolī ] sahajolis \vl}\\!}
\end{versinnote}
\end{sources}

%\begin{philcomm}[hp03_009_1]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.10}
\begin{translation}[hp03_010]
Now the great seal (\emph{mahāmudrā}):\\
{}[The yogi] should press the perineum with the heel of the left foot, hold [the foot of] the extended leg with the hands and breathe in through the mouth.
\end{translation}

\begin{sources}[hp03_010]
\emph{Amaraugha} 19
\begin{versinnote}
\tl{pādamūlena vāmena yoniṃ saṃpīḍya dakṣiṇam |\\+}
\tl{pādaṃ prasāritaṃ dhṛtvā karābhyāṃ pūrayen mukhe ||\\+}
\tl{\var{dhṛtvā ] kṛtvā \vl}\\!}
\end{versinnote}

Cf.~\emph{Amṛtasiddhi} 11.3
\begin{versinnote}
\tl{yoniṃ saṃpīḍya vāmena pādamūlena yatnataḥ |\\+}
\tl{savyaṃ prasāritaṃ pādaṃ karābhyāṃ dhārayed dṛḍham ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_010]
\emph{Haṭharatnāvalī} 2.37
\begin{versinnote}
\tl{pādamūlena vāmena yoniṃ sampīḍya dakṣiṇam |\\+}
\tl{pādaṃ prasāritaṃ kṛtvā karābhyāṃ pūrayen mukham ||\\+}
\tl{\var{mukham ] mukhe \vl}\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{pādamūlena vāmena yoniṃ saṃpīḍya dakṣiṇam |\\+}
\tl{pādaṃ prasāritaṃ dhṛtvā karābhyāṃ pūrayen mukham ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.178 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{pādamūlena vāmena yoniṃ sampīḍya dakṣiṇam |\\+}
\tl{pādaṃ prasāritaṃ kṛtvā karābhyāṃ dhārayed dṛḍham ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.11}
\begin{translation}[hp03_011]
[The yogi] should apply a lock to the throat and hold the breath in the upper [part of the body]. Just as a snake hit with a staff assumes the form of a staff, [\dots] 
\end{translation}

\begin{sources}[hp03_011]
\emph{Amaraugha} 20
\begin{versinnote}
\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_011]
\emph{Haṭharatnāvalī} 2.37cd–38ab
\begin{versinnote}
\tl{kaṇṭhe bandhaṃ samāropya pūrayed vāyum ūrdhvataḥ ||\\+}
\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} ff.~73r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.179 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{kaṇṭhe bandhaṃ samāropya dhārayed vāyum ūrdhvataḥ |\\+}
\tl{yathā daṇḍāhataḥ sarpo daṇḍākāraḥ prajāyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_011]
The instruction to hold the breath upwards (\emph{ūrdhvataḥ}) is somewhat vague. In a commentarial passage on this verse in \emph{Yuktabhavadeva} 7.187, Bhavadevamiśra clarifies this by saying, `one should hold it higher than the heart' (\emph{hṛdayād ūrdhvato dhārayet}).
\end{philcomm}

%%%%%%%%%%
\subsection*{3.12}
\begin{translation}[hp03_012]
[\dots] so the goddess \emph{kuṇḍalinī} suddenly becomes straight. Then she becomes still in the vessel with two halves.
\end{translation}

\begin{sources}[hp03_012]
\emph{Amaraugha} 21
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahasā bhavet |\\+}
\tl{tadāsau maraṇāvasthā jāyate dvipuṭāśritā ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_012]
\emph{Haṭharatnāvalī} 2.38cd–39ab
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet ||\\+}
\tl{tathā sā maraṇāvasthā jāyate dvipuṭīsthitā |\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} ff. 72v–73r (attr.~to the \emph{Haṭhapradīpikā}) % order of the hemistichs has been changed. I don't know if or where I should mention this.
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahajā bhavet |\\+}
\tl{tathāsau maraṇāvasthāṃ harate dvipuṭāśrayām ||\\+}
\tl{dvipuṭā ] em., dvipaṭā° L, dvipadā° N. \\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.180 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ṛjvībhūtā tathā śaktiḥ kuṇḍalī sahasā bhavet |\\+}
\tl{tadāsau maraṇāvasthā jāyate dvipuṭāśritā ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_012]
This verse is taken from the \emph{Amaraugha}, which uses the alchemical imagery of the \emph{Amṛtasiddhi} to describe the stilling of \emph{kuṇḍalinī} in the central channel. Drawing on Hellwig 2009:238–240, Mallinson and Szanto (2021:21) note that “In alchemical texts \emph{māraṇa} (“killing”) involves heating a substance and thereby changing its state, usually through calcination or oxidation, so that it becomes inert. In the \emph{Amṛtasiddhi māraṇa} and other derivatives of the root \emph{mṛ}, “die”, are used to denote the stilling or stopping of either the breath or Bindu.” Thus when \emph{kuṇḍalinī} is said to be in the state of \emph{maraṇa} the meaning is that she is stilled. The \emph{dvipuṭa} or “vessel with two halves” in which this occurs is the same as the \emph{Amṛtasiddhi}’s \emph{saṃpuṭa}, which, drawing on Hellwig 2009:342, Mallinson and Szanto (2021:22) say “consists of two \emph{puṭa}s joined together to form a sealed crucible for heating reagents without evaporation”. In the yoga of the \emph{Amṛtasiddhi}, the bodily \emph{saṃpuṭa} is formed by applying locks at the top and bottom of the central channel, i.e.~constricting the perineal region and the throat. In the \emph{Haṭhapradīpikā} it is formed by pressing the perineum with the heel and constricting the throat.

As Birch (2019: 971) notes, it is unlikely that later non-Buddhist authors understood \textit{maraṇāvasthā} and \emph{dvipuṭa} according to the alchemical metaphors of the \textit{Amṛtasiddhi}. Later commentators take \emph{dvipuṭa} as the two nostrils (e.g., \textit{Yuktabhavadeva} 7.187, \textit{dvināsāpuṭa}); the \textit{iḍā} and \textit{piṅgalā} channels (e.g., \textit{Jyotsnā} 3.27, \textit{puṭayor dvayam iḍāpiṅgalayor yugmam}); or the in and out flows of the breath (e.g., \textit{Yogaprakāśikā} 5.16–17, \textit{vāyor bahirnirgamanam antaḥpraveśa iti yat puṭadvayaṃ tam}). How these commentators understood \textit{asau māraṇāvasthā} is less clear. Brahmānanda seems to take it as the death of \textit{prāṇa}, or in other words, the absence of the breath, in the two nostrils (\textit{maraṇāvasthā jāyate kuṇḍalībodhe sati suṣumnāyāṃ praviṣṭe prāṇe dvayoḥ puṭayoḥ prāṇaviyogāt}).  Bhavadeva thought that \emph{kuṇḍalinī}, along with \textit{prāṇa} and \textit{apāna}, remains in the two nostrils while the breath is being held (\textit{evaṃ vāyudhāraṇāyāṃ kriyamāṇāyāṃ vyākulā bhūtā kuṇḍalinī apānaprāṇābhyāṃ saha nāsāpuṭadvayāśritā bhavati}). Others, such as Śivānanda and Bālakṛṣṇa, favour the reading \textit{tadā sā maraṇāvasthāṃ harate dvipuṭāśritām} (or \textit{dvipuṭāśrayām}), which is present in group 2 and 3 manuscripts of the \textit{Haṭhapradīpikā}. Bālakṛṣṇa understands this to mean that the great seal destroys death (\textit{maraṇāvasthāṃ harate mahāmudreti bhāvaḥ}) but it could also mean that the awakened \emph{kuṇḍalinī} destroys death, which is usually dependent on the in and out breaths.

%JB: I think our translation should reflect a likely Śaiva interpretation rather than the alchemical metaphors in the Amṛtasiddhi (e.g., `Then, that state of death arises in the two nostrils'). JM: this would imply that we think that Svātmārāma decided to use the verse with this understanding, which seems unlikely to me. JB. Unlikely? All the evidence in commentaries on the HP and later compendiums, like the YCM and Yuktabhavadeva, indicates that the alchemical metaphors of the AS were not understood here. It's not even clear that the alchemical metaphors work with raising moving or raising kuṇḍalinī. In contrast to this, there's not one commentarial explanation or rewrite of the verse that reveals an alchemical meaning. We also have to bear in mind that there's no evidence that Svātmārāma and later commentators knew the AS. In the one case where there is some evidence (i.e., Śivānanda quotes it in the YCM), it doesn't seem to have helped the author understand the alchemical metaphor in this instance (as Śivānanda rewrote the verse to say that kuṇḍalinī destroys the state of death).

%The meaning of "Śaiva" here and above is unclear to me. Doesn't it work to say  "later authors"? JB I've changed the note above to non-Buddhist.
\end{philcomm}

%%%%%%%%%%
\subsection*{3.13}
\begin{translation}[hp03_013]
{}[The yogi] should then exhale very slowly, not quickly. This is the great seal revealed by the great Siddhas.
\end{translation}

%\begin{sources}[hp03_013]
%\end{sources}

\begin{testimonia}[hp03_013]
\emph{Yuktabhavadeva} 181 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{tataḥ śanaiḥ śanair eva recayen na tu vegataḥ |\\+}
\tl{iyaṃ khalu mahāmudrā mahāsiddhaiḥ pradṛśyate ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_013]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.14}
\begin{translation}[hp03_014]
Problems such as the great afflictions [and] death and so forth are removed [by it], and that is why the most wise call it the `great seal.'
\end{translation}

\begin{sources}[hp03_014]
\emph{Amaraugha} 22
\begin{versinnote}
\tl{mahākleśādayo doṣā bhidyante maraṇādayaḥ |\\+}
\tl{mahāmudrāṃ tu tenaiva vadanti vibudhottamāḥ ||\\+}
\tl{\var{mahākleśādayo doṣā ] mahārogā mahākleśā \vl}\\+}
\tl{\var{bhidyante ] jīryante \vl}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_014]
\emph{Yogacintāmaṇi} f.~72v (attr.~to the \emph{Skandapurāṇa})
\begin{versinnote}
\tl{mahākleśā yato doṣā jīryante maraṇādayaḥ |\\+}
\tl{mahāmudrāṃ ca tenaiva vadanti vibudhottamā iti ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.182 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{mahākleśā yato doṣā jīryante maraṇādayaḥ |\\+}
\tl{mahāmudrā ca tām eva vadanti vibudhottamāḥ ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_014]
%Discuss the group 2 reading \emph{mahākleśā yato}. Also the V19 reading \emph{mahākleśā mahādoṣā}, which is similar to the \emph{Amaraughaprabodha}'s reading (\textit{mahārogā mahākleśā}). [JB: can't see a need for a note as alpha reading is close to the Amaraugha.]
%JM: note on assonance with the sounds of first line supporting the name?
This verse might be explaining the name of \emph{mahāmudrā} through assonance with \emph{mahākleśādayo, doṣā} and \emph{maraṇādayaḥ} in the first line.
\end{philcomm}

%%%%%%%%%%
\subsection*{3.15}
\begin{translation}[hp03_015]
After practising on the lunar side of the body, the yogi should then practise on the solar side. [The yogi] should finish practising the seal when the count is even.
\end{translation}

\begin{sources}[hp03_015]
\emph{Vivekamārtaṇḍa} 60
\begin{versinnote}
\tl{candrāṅge tu samabhyasya sūryāṅge punar abhyaset |\\+}
\tl{yāvat tulyā bhavet saṅkhyā tato mudrāṃ visarjayet ||\\!}
%\tl{\var{candrāṅge tu ] candrāṅgena \vl • sūryāṅge punar abhyaset ] sūryāṅgenābhyaset punaḥ \vl}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_015]
\emph{Yuktabhavadeva} 7.183 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{candrāṅgena samabhyasya sūryāṅgenābhyaset tataḥ |\\+}
\tl{yāvat tulyā bhavet saṃkhyā tato mudrāṃ visarjayet ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp03_015]
The terms \emph{candrāṅga} and \emph{sūryāṅga} are unusual and not used in other yoga texts outside the context of \emph{mahāmudrā}. In \emph{Jyotsnā} 3.15, Brahmānanda glosses \emph{candrāṅga} as \emph{vāmāṅga} (`the left side of the body') and \emph{sūryāṅga} as \emph{dakṣāṅga} (`the right side of the body') and goes on to explain the sequence of practice as follows:

\begin{quote}
This is the sequence in the [practice]. Joining the heel of the bent left leg with the region of the perineum and holding the big toe of the extended right leg with bent index fingers bent is the practice, that is, the practice on the left side of the body (\emph{vāmāṅga}). In this practice, the inhaled breath remains on the left side of the body. Joining the heel of the bent right leg with the region of the perineum and holding the big toe of the extended left leg with bent index fingers is the practice, that is, the practice on the right side of the body. In this practice, the inhaled breath remains on the right side of the body.
\emph{atrāyaṃ kramaḥ} | \emph{ākuñcitavāmapādapārṣṇiṃ yonisthāne saṃyojya prasāritadakṣiṇapādāṅguṣṭham ākuñcitatarjanībhyāṃ gṛhītvābhyāso vāmāṅge 'bhyāsaḥ} | \emph{asminn abhyāse pūrito vāyur vāmāṅge tiṣṭhati} | \emph{ākuñcita\-dakṣa\-pāda\-pārṣṇiṃ yonisthāne saṃyojya prasārita\-vāma\-pādā\-ṅguṣṭham ākuñcita\-tarjanībhyāṃ gṛhītvābhyāso dakṣāṅge 'bhyāsaḥ} | \emph{asminn abhyāse pūrito vāyur dakṣāṅge tiṣṭhati} |
\end{quote}

\end{philcomm}

%%%%%%%%%%
\subsection*{3.16}
\begin{translation}[hp03_016]
{}[For the yogi who practises thus] there is no wholesome or unwholesome [food], and all flavours without exception become flavourless. Even terrible poison, when consumed, is digested like nectar.
\end{translation}

\begin{sources}[hp03_016]
\emph{Vivekamārtaṇḍa} 61
\begin{versinnote}
\tl{na hi pathyam apathyaṃ vā rasāḥ sārve 'pi nīrasāḥ |\\+}
\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\+}
\tl{\var{jīryati] jīryate \vl, jāyate \vl}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_016]
\emph{Haṭharatnāvalī} 2.40
\begin{versinnote}
\tl{na hi pathyam apathyaṃ vā rasāḥ sarve 'pi nīrasāḥ |\\+}
\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryate ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~73r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{na hi pathyam apathyaṃ vā rasāḥ sarve 'pi nīrasāḥ |\\+}
\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 7.184 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{iha pathyam apathyaṃ vā sarasaṃ nirasaṃ ca vā |\\+}
\tl{api bhuktaṃ viṣaṃ ghoraṃ pīyūṣam iva jīryati ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp03_016]
%In \emph{Jyotsnā} 3.16, Brahmānanda understands the second verse quarter to mean all flavours –pungent (\emph{kaṭu}), acid (\emph{amlā}), etc.– and stale (\emph{nīrasa}) [food] and takes it with the verb \emph{jīryati}.
%[JB: This note doesnt seem relevant to our translation]
%\end{philcomm}

%%%%%%%%%%
\subsection*{3.17}
\begin{translation}[hp03_017]
Diseases such as consumption, skin afflictions, constipation, swelling and indigestion disappear for [the yogi] who practises the great seal.
\end{translation}

\begin{sources}[hp03_017]
\emph{Vivekamārtaṇḍa} 62
\begin{versinnote}
\tl{kṣayakuṣṭagudāvarttagulmājīrṇajvaravyathāḥ |\\+}
\tl{tasya doṣāḥ kṣayaṃ yānti mahāmudrāṃ tu yo 'bhyaset ||\\+}
\tl{\var{tasya doṣāḥ ] sarvarogāḥ \vl, rogās tasya \vl}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp03_017]
\emph{Haṭharatnāvalī} 2.41
\begin{versinnote}
\tl{kṣayakuṣṭhagudāvartagulmājīrṇapurogamāḥ |\\+}
\tl{doṣāḥ sarve kṣayaṃ yānti mahāmudrāṃ tu yo 'bhyaset ||\\!}
\end{versinnote}
